---
title: "Animating geometric optimization: small polygons"
author: "John C. Nash, A. N. Other"
date: "`r Sys.Date()`"
output: pdf_document
## output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using and extending the optimr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Roptanimation** is an experimental R package to display the progress of 
geometric animations. A classic example is the **largest small polygon** where
we aim to maximize the area of the polygon subject to the constraint that no
two vertices are separated by more than one unit of distance.

## Background

The **The Largest Small Hexagon** is the title of a paper by
Ron Graham (J. Combinatorial Theory (A), vol. 18, pp. 165-170, 1975). This did
not introduce this problem, but served to bring it to wider attention. One of
the authors (JCN) used this problem to illustrate constrained optimization using
the tools in J.C. Nash and M. Walker-Smith (1987, Nonlinear parameter estimation:
an integrated system in BASIC, now available online at 
https://archive.org/details/ost-engineering-jnmws2004).

To provide a visual presentation of the optimization, Nash coded a display for the
IBM PC family of MS DOS computers running GWBASIC. In May 2016, the discovery that
files for this example would still execute more or less satisfactorily raised the
possibility of bringing them up to date. R was a logical choice for such an 
implementation, given that the authors all work with this software system. 

## Parametrization of the polygon

For a polygon with \texttt{nv} vertices, we have \texttt{2*nv} cartesian (i.e., x, y)
coordinates. However, use of cartesian coordinates as parameters for this problem 
leads to a very complicated specification, since 3 parameters can be fixed right
away. That is, we can fix one vertex at the \texttt{(0, 0)} or origin of our [2D] space.
Moreover, we can put the second vertex at \texttt{(b[1], 0)} where \texttt{b} is a vector of
\texttt{(2*nv - 3)} parameters. Changing to a representation that uses a radius from the
origin for vertex \texttt{L} equal to \texttt{b[L-1]}, we could use the angle of this vertex from the
positive \texttt{x} axis as a parameter. Call this angle \texttt{alpha[L]}. Clearly \texttt{alpha[1]} for 
vertex 2 is 0, so the 2nd vertex is still at \texttt{(b[1], 0)}. 

We could put the alpha angles in the parameter vector as b[L+lshift] where 
lshift = nv - 3. Thus the first non-zero angle is for vertex 3 and is parameter
3 + nv - 3 = nv. Check: there are (nv - 1) radius parameters, so the first angle
parameter is in position nv. There may be good implementations based on having
parameters b[nv] ... b[2*nv-3] equal to the angles for points 2...nv. However, 
that then requires the angles to be monotonically increasing. By specifying that
b[L+lshift] = alpha[L-1] - alpha[L-2] for L=3 ... nv, we automatically get the
angles alpha monotonic by imposing a lower bound of 0 on the parameters b. 

Note that the radii cannot be negative (in fact, zero is a bad idea too), so a lower
bound of 0 can be applied to all the parameters b. An upper bound of 1 clearly applies to
the first (nv - 1) parameters. The other (nv - 2) parameters are angles in radians. If
we are to have the polygon in the positive y half-space in cartesian coordinates, 
then pi is an obvious (and likely conservative) bound on these angles. In fact,
pi is a bound on their sum.

## Problem setup

The above parametrization is implemented in the function \texttt{polysetup(nv, defsize, qpen)},
where \texttt{defsize} is the default "size" of a regular polygon for which initial parameters
are established. \texttt{qpen} is a multiplicative factor for a penalty function that may be
used to impose the distance constraints. However, the form of this penalty is not
yet given. 

```{r cache=FALSE, echo=FALSE}
library(knitr) # This is critical -- or you don't get read_chunk()
read_chunk('../R/smallpoly.R')
```


```{r polysetup}
```


## The polygon area

The parameterization of the problem allows the area to be computed as the sum of the areas of
the triangles made up from vertex \texttt{1} with vertex \texttt{L} and \texttt{(L+1)}
where \texttt{L} runs from \texttt{2} to \texttt{(nv-1)}.

```{r polyarea}
```


## Conversion of radial to cartesian coordinates

For drawing the current polygon, we need cartesian coordinates rather than the specially 
organized radial coordinates defined by the optimization parameters. The R function
\texttt{polypar2XY} carries out this computation and puts the \texttt{x, y} coordinates
in a two-vector list \texttt{XY}. \texttt{XY\$x} gives the \texttt{x} coordinates and 
\texttt{XY\$y} gives the \texttt{y} coordinates. To simplify the plotting of the polygon
the first and last values of each list are both 0 so that a graph with joining lines 
automatically gives the closed figure polygon.

```{r polypar2XY}
```

## Distance between polygon vertices

To verify constraints and to construct penalty or barrier functions for the optimization
process for this problem, we also need vertex to vertex distances. These are computed by 
the function \texttt{polydistXY}. This function uses the cartesian coordinates for the 
current polygon that result from running the function \texttt{polypar2XY}

```{r polydistXY}
```

## Testing the functions presented so far.

A very simple script calling the above functions allows us to verify that they are
working as expected. Note that for this script to work properly, we need to ensure
that the graph is drawn on a square grid. 

```{r polyex1, fig.width=5, fig.height=5}
```
