---
title: "River basin pollution game"
author: "Christophe Dutang, Clémentine Prieur"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: yes
    number_sections: yes
    includes:
      in_header: "GNE-preamble.tex"
    citation_package: natbib  
bibliography: "GNE.bib"
biblio-style: plainnat
geometry: margin=2cm
documentclass: article
classoption: a4paper
---
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 8)
```



# Game description 

We consider the River Basin Pollution game proposed by \cite{haurie1997optimal} 
and studied by \cite{krawuryasev00} and \cite{krawczyk05}.
$J$ players are engaged in an economic activity (e.g. paper pulp producing) along a river.
\cite{contreras2018playing} consider a similar approach for thermal electricity 
generators.


## Setting of the game

The pollution game proposed in this paper combines a linear approximation of the 
time-invariant steady state of pollution PDE dynamics and
a non-seasonal differential game proposed by \cite[Section 2]{haurie1997optimal}. 
Authors argue that ``The river water's dynamics is usually faster than the underlying economic process.
This implies that, even if the economic planning horizon is relatively short (e.g. years), 
from the river model's perspective, the market instruments (e.g. taxes)
will be applied over a very long (almost infinite) period.
This observation permits us to concentrate on the time-invariant 
steady state solution of the partial differential
equation which we are using to model the fundamental dynamics of the river stream, 
that receives pollution from
infinitely many sources, transports and dilutes it.''
This results in solving a static non-cooperative game with coupled constraints based on the
steady-state pollution concentration level. 



The players choose the economic level/output $(x_j)_j$ while meeting environmental conditions set
by local authorities.
However, pollutants may be expelled into the river, where they disperse. 
The joint production externatilities of all players must be satisfied at monitoring stations along
the river.
The pollution is expelled into the river where it disperse, decays, reaches monitoring stations.
The monitoring stations $l = l_1, l_2,\dots$ located along the river are used by the local authority 
to set maximum pollutant concentration levels 
$$
q_l(\bx) = \sum_{j=1}^J u_{j,l} e_j x_j \leq K_l.
$$
That is, we assume that pollutant is linearly $e_j x_j$ by agent's output $x_j$
with emission rate $e_j$ and $u_{j,l}$ are decay-transportation coefficient 
of pollution from player $j$ at control station $l$.
Emission rate $e_j$ represents the industrial effort to comply with pollution standard
while the decay coefficient $u_{j,l}$ is linked to the distance between the industrial $j$
and the quality test station at $l$.

\cite{haurie1997optimal} use the following components $R_j$, $F_j$ to build the objective function $O_j$.
The revenue of player $j$ is 
$$
R_j(\bx) = (d_{1} - d_{2} \times (x_1+\dots+x_J))\times x_j,
$$
The associated cost of player $j$ is
$$
F_j(\bx) = (c_{1,j} + c_{2,j} x_j) \times x_j.
$$
The net profit is thus
$$
O_j(\bx) = (d_1 - d_2 \times (x_1+\dots+x_J) - c_{1,j} - c_{2,j} x_j)\times x_j.
$$
The game consists in finding solutions (generalized Nash equilibrium) of 
$$
\forall j, \max_{x_j} O_j(\bx), \text{such that } \forall l, q_l(\bx) \leq K_l.
$$
Parameters are listed in  Table \ref{tab:param:signif}. 


## Variables and parameters

\cite{haurie1997optimal} subsequent works by \cite{krawuryasev00}, \cite{krawczyk05},
\cite{contreras2018playing} always examplify their work on a three-player two-station
game with parameters given in Table \ref{tab:param:value}.
See also Figure \ref{fig:riverline} for a descriptive diagram.


\begin{table}[htb]
\centering
\begin{tabular}{cc|cc}
\multicolumn{2}{c}{Notation of \cite{krawuryasev00}, \cite{krawczyk05}}
&
\multicolumn{2}{c}{Notation of \cite{haurie1997optimal}} \\
\hline
$x_j$ & utilisation level & $x_j$\\
$d_{j,1}, d_{j,2}$ & demand levels & $\psi_{j,1}$,  $\psi_{j,2}$ \\
$x_j\leq y_j$ & installed capacity & $x_j\leq y_j$ & installed capacity \\
$u_{j,l}$ & decay/transportation coefficients & $\alpha_{j,l}$ & physical investment    \\
$c_{l,j}$ & cost  & $\varphi_{l,j}$ \\
$e_j\in[0,1]$ & emission coefficient & $emra_\theta$ & emission rates \\
$K_l$ & pollution constraints & $q_{max,l}$ \\
\hline
\end{tabular}
\caption{Parameter details}
\label{tab:param:signif}
\end{table}

\begin{table}[htb]
\centering
\begin{tabular}{ccc|c|cc}
Player $j$ & $c_{1,j}$ & $c_{2,j}$ & $e_{j}$ & $u_{1,j}$ & $u_{2,j}$   \\
\hline
1 & 0.1 & 0.01 & 0.5 & 6.5 & 4.583   \\
2 & 0.12 & 0.05 & 0.25 & 5 & 6.25   \\
3 & 0.15 & 0.01 & 0.75 & 5.5 & 3.75  \\
\hline
Other & $d_{1}$ & $d_{2}$ & $K_1$ & $K_2$ \\
\hline
& 3 & 0.01 & 100 & 100 \\
\hline
\end{tabular}
\caption{Parameter values of \cite{krawuryasev00} and others}
\label{tab:param:value}
\end{table}


\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{img/river-basin-pollution-game-indus-only}
  \caption{Diagram of the river line}
  \label{fig:riverline}
\end{figure}



<!-- \begin{tabular}{lc} -->
<!-- $reca_j$ & residual capacity \\ -->
<!-- $life_\theta$ & life duration \\ -->
<!-- $maco_\theta$ & unit maintenance cost \\ -->
<!-- $inco_\theta$ & unit investment cost \\ -->
<!-- $usco_\theta$ & unit utilisation cost \\ -->
<!-- $cont_\theta$ & contribution output \\ -->
<!-- \end{tabular} -->




## Purpose of the game

\cite{krawczyk05} computes the GNE $\bx^\star$ and observe that only the first constraint
(at location 1 in the river) is active $q_1(\bx^\star)=K_1$ while the second constraint is in active
$q_2(\bx^\star)<K_2$.

\cite{krawczyk05} and \cite{krawczyk2007numerical} 
consider the cooperative/social version of the game where the objective
is $\sum_j O_j(\bx)$ in order to compare with the competitive equilibrium.
They got a cooperative equilibrium $\hat\bx$ with a higher objective aggregate than
the aggregate of optimal objectives for the competitive equilibrium $\bx^\star$.


\cite{krawczyk05} and \cite{krawczyk2007numerical} also investigates the use of a 
self-enforcing Pigouvian tax which penalizes each player.
His numerical illustrations exhibit an equivalent equilibrium between the original
constrained game and the game with tax-penalized objective functions.

\cite{elnaboulsi2011efficient} also consider the unconstrained version of 
the pollution game, for which the optimal utilization level $x_j$ and associated
emission is far higher the constrained version, see their Table 2.

# Sensitivity analysis 

## Base run

The constrained equilibrium is given in \cite{krawczyk05} and \cite{krawczyk2007numerical}.
Tables below give what we obtain by using the GNE package.

```{r}
#Rosen algorithm
opteq <- c(21.149, 16.028, 2.722)
optprofit <- c(48.42, 26.92, 6.60)
optconstr <- c(100, 81.17)
lagrmult <- c(0.574, 0)
```

```{r}
require(GNE)
#-------------------------------------------------------------------------------
# (3) River basin pollution game of Krawczyk and Stanislav Uryasev (2000)
#-------------------------------------------------------------------------------

myarg0 <- list(
  C = cbind(c(.1, .12, .15), c(.01, .05, .01)),
  U = cbind(c(6.5, 5, 5.5), c(4.583, 6.25, 3.75)),
  K = c(100, 100),
  E = c(.5, .25, .75),
  D = c(3, .01)
)

dimx <- c(1, 1, 1)
#O_i(x)
obj <- function(x, j, arg)
{
  (arg$D[1] - arg$D[2]*sum(x[1:3]) - arg$C[j, 1] - arg$C[j, 2]*x[j]) * x[j]
}
#Gr_x_j O_i(x)
grobj <- function(x, i, j, arg)
{
  dij <- 1*(i == j)
  res <- -(-arg$D[2] - arg$C[i, 2]*dij) * x[i] 
  res - (arg$D[1] - arg$D[2]*sum(x[1:3]) - arg$C[i, 1] - arg$C[i, 2]*x[i]) * dij
}
#Gr_x_k Gr_x_j O_i(x)
heobj <- function(x, i, j, k, arg)
{
  dij <- 1*(i == j)
  dik <- 1*(i == k)
  
  arg$D[2] * dik + arg$D[2] * dij + 2 * arg$C[i, 2] * dij * dik
}

dimmu <- 5
  #h(x)
  h <- function(x, arg)
    c(sum(arg$U[, 1] * arg$E * x[1:3]) - arg$K[1],
      sum(arg$U[, 2] * arg$E * x[1:3]) - arg$K[2],
      -x[1],
      -x[2],
      -x[3])
  #Gr_x_j h(x)
  grh <- function(x, j, arg)
    c(arg$U[j, 1] * arg$E[j], 
      arg$U[j, 2] * arg$E[j], 
      -1*(1 ==j), 
      -1*(2 ==j), 
      -1*(3 ==j))
  #Gr_x_k Gr_x_j g_i(x)
  heh <- function(x, j, k, arg)
    c(0, 0, 0, 0, 0)
  

#true value around (21.146, 16.027, 2.724, 0.574, 0.000)
z0 <- rep(1, sum(dimx)+sum(dimmu))

getNE <- function(x, control=list(maxit=100, trace=0), check=TRUE, output="equilibrium")
{
  res <- sapply(1:NROW(x), function(i)
    {
      myarg <- list(
        C = cbind(x[i,paste("C",1:3,sep="")], x[i,paste("C",1:3+3,sep="")]),
        U = cbind(x[i,paste("U",1:3,sep="")], x[i,paste("U",1:3+3,sep="")]),
        K = x[i,paste("K",1:2,sep="")],
        E = x[i,paste("E",1:3,sep="")],
        D = x[i,paste("D",1:2,sep="")]
      )
      z0 <- rep(1, sum(dimx)+sum(dimmu))
      
      res <- GNE.nseq(z0, dimx, dimmu=dimmu, grobj=grobj, arggrobj=myarg, heobj=heobj, argheobj=myarg, 
                      joint=h, argjoint=myarg, grjoint=grh, arggrjoint=myarg, hejoint=heh, arghejoint=myarg, 
                      compl=phiFB, gcompla=GrAphiFB, gcomplb=GrBphiFB, method="Newton", 
                      control=control)
      
      if(output == "equilibrium")
        return(res$par[1:3])
      else
        return(res)
    }
    )
    t(res)
}

basesetting <- matrix(unlist(myarg0), nrow=1, dimnames = list(NULL, names(unlist(myarg0))))
resbase <- getNE(basesetting, output="full")
```
```{r, echo=FALSE}
knitr::kable(cbind("Numerical"=resbase[,1]$par[1:3], "Theoretical"=opteq), caption="equilibrium value")

resprofit <- sapply(1:3, function(j) 
  obj(resbase[,1]$par, j, myarg0))
knitr::kable(cbind("Numerical"=resprofit, "Theoretical"=optprofit), caption="equilibrium profit")

resconstr <- h(resbase[,1]$par, myarg0)
knitr::kable(cbind("Numerical"=myarg0$K + resconstr[1:2], "Theoretical"=optconstr), caption="equilibrium constraints")

knitr::kable(cbind("Numerical"=resbase[,1]$par[4:5], "Theoretical"=lagrmult), caption="Lagrange multipliers")
```


## Generation of parameter distributions

Parameters listed in Tables \ref{tab:param:signif} and \ref{tab:param:value} obey to a logical
interpretation. 
Overall parameters $d_i$'s should be thought by pair.
We assume $d_1,d_2$ follow a bivariate normal distribution with a 50\%-correlation,
and mean centered at the base value.

Other parameters $c_{i,j}$'s act on the cost terms of player $j$ and are correlated.
We assume $c_{1,1},\dots,c_{2,3}$ follow a 6-dimensional normal distribution with a 50\%-correlation,
and mean centered at the base value.


Finally constraint parameters $e_j$, $u_{j,l}$ and $K_i$'s act together on the game.
Due to the multiplicative effect of $e_j$, $u_{j,l}$, we only consider the effect of emission rates $e_j$.
We assume emission rate to be uniformly distributed in $(0.1,0.9)$ while $K_1$, $K_2$
follow a bivariate normal distribution with a 50\%-correlation,
and mean centered at the base value.

```{r}
par(mfrow=c(1,1), mar=c(4,4,2,1))

gend1d2 <- function(n, d1d2.mean, d1d2.rho, d1d2.sig)
{
  require(mvtnorm)
  Sig <- diag(d1d2.sig) %*% cbind(c(1, d1d2.rho), c(d1d2.rho, 1)) %*% diag(d1d2.sig)
  #print(Sig)
  res <- rmvnorm(n, d1d2.mean, Sig)
  #print(res)
  res <- pmax(res, 0.005)
  colnames(res) <- c("D1", "D2")
  res
}
pairs(gend1d2(1000, myarg0$D, 0.5, c(0.10, 0.025)))

gencij <- function(n, cij.mean, cij.rho, cij.sig)
{
  require(mvtnorm)
  R <- matrix(cij.rho, 6, 6)
  diag(R) <- 1
  #print(R)
  Sig <- diag(cij.sig) %*% R %*% diag(cij.sig)
  #print(Sig)
  res <- rmvnorm(n, cij.mean, Sig)
  #print(res)
  res <- pmax(res, 0.005)
  #lapply(1:NROW(res), function(i)
  #{
  #  xi <- matrix(res[i,], nrow=3)
  #  colnames(xi) <- c("c1j", "c2j")
  #  xi
  #})
  colnames(res) <- paste0("C", 1:6)
  res
}
gencij(3, as.vector(myarg0$C), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01))
pairs(gencij(1e3, as.vector(myarg0$C), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01)))

getej <- function(n, low, upp)
{
  res <- cbind(runif(n, low, upp), runif(n, low, upp), runif(n, low, upp))
  colnames(res) <- paste0("E", 1:3)
  res
}
pairs(getej(1e3, .1, .9))


getUj <- function(n, Uj.mean, Uj.rho, Uj.sig)
{
  require(mvtnorm)
  R <- matrix(Uj.rho, 6, 6)
  diag(R) <- 1
  #print(R)
  Sig <- diag(Uj.sig) %*% R %*% diag(Uj.sig)
  #print(Sig)
  res <- rmvnorm(n, Uj.mean, Sig)
  #print(res)
  res <- pmax(res, 0.005)
  colnames(res) <- paste0("U", 1:6)
  res
}

pairs(getUj(1e3, as.vector(myarg0$U), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01)))

getKj <- function(n, Kj.mean, Kj.rho, Kj.sig)
{
  require(mvtnorm)
  Sig <- diag(Kj.sig) %*% cbind(c(1, Kj.rho), c(Kj.rho, 1)) %*% diag(Kj.sig)
  #print(Sig)
  res <- rmvnorm(n, Kj.mean, Sig)
  #print(res)
  res <- pmax(res, 0.005)
  colnames(res) <- c("K1", "K2")
  res
}

pairs(getKj(1e3, myarg0$K, 0.5, c(5, 5)))

genXparam <- function(n, cij.mean, cij.rho, cij.sig, Uj.mean, Uj.rho, Uj.sig, 
                      Kj.mean, Kj.rho, Kj.sig, 
                      low, upp, d1d2.mean, d1d2.rho, d1d2.sig)
{
  P1 <- gencij(n, cij.mean, cij.rho, cij.sig)
  P2 <- getUj(n, Uj.mean, Uj.rho, Uj.sig)
  P3 <- getKj(n, Kj.mean, Kj.rho, Kj.sig)
  P4 <- getej(n, low, upp)
  P5 <- gend1d2(n, d1d2.mean, d1d2.rho, d1d2.sig)
  cbind(P1, P2, P3, P4, P5)
}

cstU <- as.vector(myarg0$U)
names(cstU) <- paste0("U", 1:6)

#test
X1 <- genXparam(10, as.vector(myarg0$C), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          as.vector(myarg0$U), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          myarg0$K, 0.5, c(5, 5), .1, .9,
          myarg0$D, 0.5, c(0.10, 0.025)) 

fX1 <- getNE(X1[1:10, ])
dim(fX1)

n <- 1e2
X1 <- genXparam(n, as.vector(myarg0$C), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          as.vector(myarg0$U), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          myarg0$K, 0.5, c(5, 5), .1, .9,
          myarg0$D, 0.5, c(0.10, 0.025)) 
dim(X1)
X2 <- genXparam(n, as.vector(myarg0$C), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          as.vector(myarg0$U), 0.5, c(0.1, 0.1, 0.1, 0.01, 0.01, 0.01),
          myarg0$K, 0.5, c(5, 5), .1, .9,
          myarg0$D, 0.5, c(0.10, 0.025)) 
dim(X2)

```



## Sensitivity analysis of multivariate outputs

```{r}
library(sensitivity)
multSob1 <- sobolMultOut(getNE, q=3, X1 = X1, X2 = X2,
                         MCmethod = "sobol")
multSob1
plot(multSob1, ylim=range(multSob1$S))
grid()

```


## Sensitivity analysis of univariate outputs


### Arithmetic mean of optimal objectives $O_j(\bx^\star)$

```{r}
Youtputs <- getNE(X1)

ObjOutputs <- sapply(1:3, function(j)
  sapply(1:NROW(Youtputs), function(i)
  obj(Youtputs[i,], j, myarg0)
  )
)

Objarithmean <- apply(ObjOutputs, 1, mean)

(respccArith <- pcc(data.frame(X1), Objarithmean))

(ressrcArith <- src(data.frame(X1), Objarithmean))

plot(respccArith)
grid()
plot(ressrcArith)
grid()

```



### Arithmetic mean of equilibrium $\bx^\star$

```{r}

Yarithmean <- apply(Youtputs, 1, mean)

(respccArith <- pcc(data.frame(X1), Yarithmean))

(ressrcArith <- src(data.frame(X1), Yarithmean))

plot(respccArith)
grid()
plot(ressrcArith)
grid()
```


### Geometric mean of equilibrium $\bx^\star$

```{r}

Ygeomean <- apply(Youtputs, 1, function(x) prod(pmax(x,0)^(1/3)))

(respccGeo <- pcc(data.frame(X1), Ygeomean))

(ressrcGeo <- src(data.frame(X1), Ygeomean))
  
plot(respccGeo)
grid()
```




# Conclusion

